---
title: "Solving sudoku's with R and C++"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is a sudoku

A sudoku is a puzzle where given a $9 \times 9$ grid of cells. A cell can be
filled with a number or be empty. The goal is to fill all the empty cells 
according to a set of rules:

- Cells may only be filled by the numbers 1 to 9.
- A number may appear only once in each row.
- A number may appear only once in each column.
- The grid is divided in a $3 \times 3$ grid of $3 \times 3$ cells each. In each 
smaller grid a number may only appear once.

## Solving Sudoko's in R and C++

In this presentation three approaches are compared to solve sudoku's:

- R
- C++
- Hybrid

## Representing a sudoku

A sudoku represented by a space separated numbers in 9 rows and 9 columns. 

For example (taken from Wikipedia):

```{r, echo = TRUE}
sudokuTxt <- "
1 0 0  0 0 0  0 0 6
0 0 6  0 2 0  7 0 0
7 8 9  4 5 0  1 0 3
0 0 0  8 0 7  0 0 4
0 0 0  0 3 0  0 0 0
0 9 0  0 0 4  2 0 1
3 1 2  9 7 0  0 4 0
0 4 0  0 1 2  0 7 8
9 0 8  0 0 0  0 0 0"
```

0 signifies an empty cell.

## Loading the sudoku

Reading in the sudoku and storing it in a matrix using R is simple:

```{r, echo = TRUE}
sudoku <- as.matrix(
  read.table(text = sudokuTxt, 
             col.names = letters[1:9]))
```

In C++ this is still easy but not quite as simple as the above.

## General solution strategy

One way to solve sudoku's is to elimation and when in doubt trying. 

- Elimination is the process of finding possible values for empty cells given
the currently known values.

- After the elimination step it may happen that empty cells remain for which
more than value is possible. In this case, the algorithm proceeds by making a 
`random' choice and continues until it finds a complete solution or a 
contradiction.
- A contradiction is a partial solution in which for some cell there is no
possible value.

## In R code

```{r, echo = FALSE}
library(zeallot)
library(magrittr)
```
```{r, echo = TRUE}
solve <- function(partialSolution, choicesFUN) {
  # Eliminate impossible values, give some suggestions
  # and flag contradictions.
  c(partialSolution, suggestions, contradiction) %<-% 
    eliminate(partialSolution, choicesFUN)
  # If dead end FALSE to trace back, if finshed TRUE.
  if (contradiction) return(list(FALSE, NULL))
  if (all(partialSolution %in% 1:9))
    return(list(TRUE, partialSolution))
  # Branching, exit when the solution is found.
  for (suggestion in suggestions) {
    c(result, solution) %<-% solve(suggestion, choicesFUN)
    if (result) return(list(result, solution))
  }
  list(FALSE, NULL)
}
```

## R Implementation - `eliminate()`

The function `eliminate()` checks for every empty cell which values are 
possible, returns a contradiction when no possible values for a cell are found
and keeps track of the cell with the least possibilities which can be used as a
pivot.

## R Implementation - `eliminate()`

```{r, echo = TRUE}
eliminate <- function(grid, choicesFUN) {
  suggestions <- 0:9
  for (i in 1:nrow(grid)) { for (j in 1:ncol(grid)) {
    if (grid[i, j] == 0L) {
      choices <- choicesFUN(grid, i, j)
      if (length(choices) == 0L) {
        return(list(NULL, NULL, TRUE))
      } else if (length(choices) == 1L) {
        grid[i, j] <- choices
        return(list(grid, list(grid), FALSE))
      } else
        suggestions <- updateSuggestions(
          choices, grid, i, j, suggestions)
    }
  }}
  list(grid, suggestions, FALSE)
}
```

## R Implementation - Helpers

```{r, echo = TRUE}
# Find the choices allowed by the rules.
findChoices <- function(grid, i, j) {
  1:9 %>% setdiff(grid[i, ]) %>%
    setdiff(grid[ , j]) %>%
    setdiff(grid[i - (i - 1) %% 3L + 0:2, 
                 j - (j - 1) %% 3L + 0:2])
}
# Create a new grid with suggested next moves.
updateSuggestions <- function(choices, grid, i, j,
                              lastBest) {
  if (length(choices) < length(lastBest))
    lapply(choices, function(choice) {
      grid[i, j] <- choice; grid
    })
  else 
    lastBest
}
```

## Solving sudoku's

With this code solving a sudoko is simple:

```{r, echo = TRUE}
# solution <- solve(sudoku)
# if (!solution[[1]]) { cat('Solution not found\n')
# } else { print(as.data.frame(solution[[2]])) }
```

## Solving harder sudoku's

This is the ``World's hardest sudoku'' (needs source)

```{r, echo = TRUE}
sudokuTxt <- "
8 0 0  0 0 0  0 0 0
0 0 3  6 0 0  0 0 0
0 7 0  0 9 0  2 0 0
0 5 0  0 0 7  0 0 0
0 0 0  0 4 5  7 0 0
0 0 0  1 0 0  0 3 0
0 0 1  0 0 0  0 6 8
0 0 8  5 0 0  0 1 0
0 9 0  0 0 0  4 0 0"
sudoku <- as.matrix(
  read.table(text = sudokuTxt,
             col.names = letters[1:9]))
```

## Solution to the ``World's hardest sudoku''

```{r, echo = TRUE}
# solution <- solve(sudoku)
# if (!solution[[1]]) { cat('Solution not found\n')
# } else { print(as.data.frame(solution[[2]])) }
```

## Some benchmarking

Benchmarking not so easy due to recursion.

## Some benchmarking
```{r, echo = TRUE}
findChoices2 <- function(grid, i, j) {
  setdiff(
    setdiff(setdiff(1:9,
                    grid[i, ]),
            grid[ , j]),
    grid[i - (i - 1) %% 3L + 0:2,
         j - (j - 1) %% 3L + 0:2])
}
```
